Deep mutational scanning of hemagglutinin helps distinguish the evolutionary fates of human H3N2 influenza virus lineages

### 2018-01-29 paper/manuscript.tex
Added baltic mention.
Minor syntax and tense edits.

\subsection*{Our measurements are consistent with existing knowledge about HA's evolution and function}
How do the HA amino-acid preferences measured in our experiments relate to the evolution of H3N2 influenza virus in nature?
This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describe H3N2 evolution compared to standard phylogenetic substitution models~\cite{bloom2017identification,hilton2017phydms}.
Table~\ref{tab:phydms} shows that the ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
This result indicates that our experiments authentically capture some of the constraints on HA evolution. 
The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}).
However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid constraints measured in our experiments (the $\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that the deep mutational scanning captures much of the purifying selection on HA.
ExpCM fit a stringency parameter that relates the selection in the experiments to that in nature~\cite{bloom2017identification,hilton2017phydms}.
The stringency parameter for our HA ExpCM is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments, but with greater stringency.
Throughout the rest of this paper, we use experimental measurements re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
These re-scaled preferences are shown in Figure~\ref{fig:logoplot}.

@@

Our experiments were performed on the Perth/2009 HA, but we are scoring mutations on a phylogenetic tree that extends from 1968 to 2017.
Because the effects of mutations can vary with genetic background~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative}, it is possible that the effects of mutations that we measured in the Perth/2009 HA might have shifted somewhat in other related HAs.
To explore this question, we scored the complete HA sequence $\mathbf{s}$ of every node in the phylogenetic tree by quantifying its average per-site sequence preference as 
\begin{equation}
F\left(\mathbf{s}\right) = \frac{1}{L}\displaystyle\sum_{r=1}^L \ln \pi_{r, s_r},
\end{equation}
where $\pi_{r, s_r}$ is the preference measured in our experiments (e.g., Figure~\ref{fig:logoplot}) for the amino acid $s_r$ at site $r$ in HA sequence $\mathbf{s}$, and $L$ is the length of the sequence.
Figure~\ref{fig:trunkvssidebranch}C shows that the sequences of trunk nodes tend to have higher preference than those of side branch nodes across the entire timespan, consistent with the finding that trunk mutations are generally more favorable than side branch mutations. 
However, the average per-site sequence preference increases as the nodes approach the Perth/2009 strain.
The Perth/2009 itself has among the highest per-site sequence preference of the entire tree, despite falling on a side branch.
We do not think that this result means that Perth/2009 is actually higher fitness than the other strains.
Rather, if there are modest shifts in the effects of mutations as HA evolves, then sequences more distant from Perth/2009 could be scored as having lower preference simply because these shifts start to degrade the accuracy of our measurements.
But importantly, our experiments consistently show the trunk to be more favorable than the side branches regardless of whether we look at sequences that precede Perth/2009 (most of Figure~\ref{fig:trunkvssidebranch}C) or sequences that occur after Perth/2009 has diverged from the trunk (Figure~\ref{suppfig:future}). 
Therefore, our experimental measurements consistently reveal the selective advantage of the trunk over more than a half-century of viral evolution.

@@

An obvious hypothesis for why the H1 deep mutational scanning is not informative about the evolutionary fate of H3 strains is that the effect of the same mutation will often be very different between these two HA subtypes.
To determine if this is the case, we can examine how much the amino-acid preferences of homologous sites have shifted between H3 and H1 HAs.
Prior experiments have found only modest shifts in amino-acid preferences between two variants of influenza nucleoprotein with 94\% amino-acid identity~\cite{doud2015site} and variants of HIV envelope (Env) with 86\% amino-acid identity~\cite{haddox2017mapping}.
However, the H1 and H3 HAs are far more diverged, with only 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A).
One simple way to investigate the extent of shifts in amino-acid preferences is to correlate measurements from independent deep mutational scanning replicates on H1 and H3 HAs.
Figure~\ref{fig:distance_distribution}B shows that replicate measurements on the same HA variant are more correlated than those on different HA variants.

@@

Despite their high sequence divergence, H1 and H3 adopt very similar protein folds~\cite{ha2002h5,russell2004h1}.
However, there are differences in the rotation and upward translation of the globular head subdomains relative to the central stalk domain among different HA subtypes~\cite{ha2002h5,russell2004h1}.
Previous work has defined clades of structurally related HA subtypes~\cite{ha2002h5,russell2004h1}. 
One such clade includes H1, H2, H5, and H6, whereas another clade includes H3, H4, and H14 HAs (Figure~\ref{fig:distance_distribution}).
Sites that are conserved at different amino-acid identities in these two clades tend to have exceptionally large shifts in amino acid preferences (Figure~\ref{fig:RMSD_heatmap}B).
The clade containing H1 has an upward shift of the globular head relative to the clade containing H3.
This structural shift has been attributed largely to the interaction between sites 107 and 75(HA2)~\cite{ha2002h5,russell2004h1}.
Specifically, the clade containing H1 has a taller turn in the interhelical loop connecting helix A and helix B in the stalk domain, and this tall turn is stabilized by a hydrogen bond between Glu-107 and Lys-75(HA2) (Figure~\ref{fig:RMSD_heatmap}C).
In deep mutational scanning of the H1 HA, site 107 has a high preference for Glu and 75(HA2) strongly prefers positively charged Lys and Arg.
In contrast, the interhelical loop in H3 HA makes a sharper and shorter turn which is facilitated by a Gly at 75(HA2).
In the deep mutational scanning of the Perth/2009 H3 HA, site 75(HA2) prefers Gly and to a lesser extent Val, while site 107 is fairly tolerant of mutations.
Therefore, some of the shifts in HA amino-acid preferences can be directly rationalized in terms of changes in HA structure.

@@

\section*{Discussion}
\label{sec:discussion}
We have measured the effect of all possible single amino-acid mutations to the Perth/2009 H3 HA on viral growth in cell culture, and demonstrated that these measurements are informative for distinguishing the evolutionary fate of human H3N2 viral strains in nature.
Specifically, HA sequences on the trunk of the phylogenetic tree tend to have mutations that our experiments measure to be more beneficial for viral growth than mutations that occur in side-branch HA sequences.
The fact that our experiments can help distinguish between trunk and side-branch lineages suggests that they might inform evolutionary forecasting.
In their landmark paper introducing predictive viral fitness models that accounted for both antigenic and non-antigenic mutations, {\L}uksza and L{\"a}ssig~\cite{luksza2014predictive} noted that the models could in principle be improved by integrating ``diverse genotypic and phenotypic data'' that more realistically represented the effects of specific mutations.
Our work suggests that deep mutational scanning can provide such data.

@@

To parse trunk mutations from side branch mutations, we defined a set of recent nodes sampled on or after January 1, 2017, and traced these nodes back to their most recent common ancestor.
We defined all branches ancestral to this most recent common ancestor as the trunk.
All nodes that do not have descendants isolated after 2017 were defined as side branch nodes.
Nodes descended from the most recent common ancestor were defined as unresolved, as it is still unclear whether these nodes will ultimately be on the trunk or side branches.
The phylogenetic tree is available as a JSON file on \comment{link to repository}.

>>>>

\subsection*{Our measurements are consistent with existing knowledge about HA's evolution and function}
How do the HA amino-acid preferences measured in our experiments relate to the evolution of H3N2 influenza virus in nature?
This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describes H3N2 evolution compared to standard phylogenetic substitution models~\cite{bloom2017identification,hilton2017phydms}.
Table~\ref{tab:phydms} shows that the ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
This result indicates that our experiments authentically capture some of the constraints on HA evolution. 
The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}).
However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid constraints measured in our experiments (the $\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that the deep mutational scanning captures much of the purifying selection on HA.
ExpCM fit a stringency parameter that relates the selection in the experiments to that in nature~\cite{bloom2017identification,hilton2017phydms}.
The stringency parameter for our HA ExpCM is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments, but with greater stringency.
Throughout the rest of this paper, we use experimental measurements re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
These re-scaled preferences are shown in Figure~\ref{fig:logoplot}.

@@

Our experiments were performed on the Perth/2009 HA, but we scored mutations on a phylogenetic tree that extends from 1968 to 2017.
Because the effects of mutations can vary with genetic background~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative}, it is possible that the effects of mutations that we measured in the Perth/2009 HA might have shifted somewhat in other related HAs.
To explore this question, we scored the complete HA sequence $\mathbf{s}$ of every node in the phylogenetic tree by quantifying its average per-site sequence preference as 
\begin{equation}
F\left(\mathbf{s}\right) = \frac{1}{L}\displaystyle\sum_{r=1}^L \ln \pi_{r, s_r},
\end{equation}
where $\pi_{r, s_r}$ is the preference measured in our experiments (e.g., Figure~\ref{fig:logoplot}) for the amino acid $s_r$ at site $r$ in HA sequence $\mathbf{s}$, and $L$ is the length of the sequence.
Figure~\ref{fig:trunkvssidebranch}C shows that the sequences of trunk nodes tend to have higher preference than those of side branch nodes across the entire timespan, consistent with the finding that trunk mutations are generally more favorable than side branch mutations. 
However, the average per-site sequence preference increases as the nodes approach the Perth/2009 strain.
The Perth/2009 itself has among the highest per-site sequence preference of the entire tree, despite falling on a side branch.
We do not think that this result means that Perth/2009 is actually higher fitness than the other strains.
Rather, if there are modest shifts in the effects of mutations as HA evolves, then sequences more distant from Perth/2009 could be scored as having lower preference simply because these shifts start to degrade the accuracy of our measurements.
But importantly, our experiments consistently show the trunk to be more favorable than the side branches regardless of whether we look at sequences that precede Perth/2009 (most of Figure~\ref{fig:trunkvssidebranch}C) or sequences that occur after Perth/2009 has diverged from the trunk (Figure~\ref{suppfig:future}). 
Therefore, our experimental measurements consistently reveal the selective advantage of the trunk over more than a half-century of viral evolution.

@@

An obvious hypothesis for why the H1 deep mutational scanning is not informative about the evolutionary fate of H3 strains is that the effect of the same mutation will often be very different between these two HA subtypes.
To determine if this is the case, we examined how much the amino-acid preferences of homologous sites have shifted between H3 and H1 HAs.
Prior experiments have found only modest shifts in amino-acid preferences between two variants of influenza nucleoprotein with 94\% amino-acid identity~\cite{doud2015site} and variants of HIV envelope (Env) with 86\% amino-acid identity~\cite{haddox2017mapping}.
However, the H1 and H3 HAs are far more diverged, with only 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A).
One simple way to investigate the extent of shifts in amino-acid preferences is to correlate measurements from independent deep mutational scanning replicates on H1 and H3 HAs.
Figure~\ref{fig:distance_distribution}B shows that replicate measurements on the same HA variant are more correlated than those on different HA variants.

@@

Despite their high sequence divergence, H1 and H3 adopt very similar protein folds~\cite{ha2002h5,russell2004h1}.
However, there are differences in the rotation and upward translation of the globular head subdomains relative to the central stalk domain among different HA subtypes~\cite{ha2002h5,russell2004h1}.
Previous work has defined clades of structurally related HA subtypes~\cite{ha2002h5,russell2004h1}.
One such clade includes H1, H2, H5, and H6, whereas another clade includes H3, H4, and H14 HAs (Figure~\ref{fig:distance_distribution}A).
Sites that are conserved at different amino-acid identities in these two clades tend to have exceptionally large shifts in amino acid preferences (Figure~\ref{fig:RMSD_heatmap}B).
The clade containing H1 has an upward shift of the globular head relative to the clade containing H3.
This structural shift has been attributed largely to the interaction between sites 107 and 75(HA2)~\cite{ha2002h5,russell2004h1}.
Specifically, the clade containing H1 has a taller turn in the interhelical loop connecting helix A and helix B in the stalk domain, and this tall turn is stabilized by a hydrogen bond between Glu-107 and Lys-75(HA2) (Figure~\ref{fig:RMSD_heatmap}C).
In deep mutational scanning of the H1 HA, site 107 has a high preference for Glu and 75(HA2) strongly prefers positively charged Lys and Arg.
In contrast, the interhelical loop in H3 HA makes a sharper and shorter turn which is facilitated by a Gly at 75(HA2).
In the deep mutational scanning of the Perth/2009 H3 HA, site 75(HA2) prefers Gly and to a lesser extent Val, while site 107 is fairly tolerant of mutations.
Therefore, some of the shifts in HA amino-acid preferences can be directly rationalized in terms of changes in HA structure.

@@

\section*{Discussion}
\label{sec:discussion}
We have measured the effect of all possible single amino-acid mutations to the Perth/2009 H3 HA on viral growth in cell culture and demonstrated that these measurements are informative for distinguishing the evolutionary fate of human H3N2 viral strains in nature.
Specifically, HA sequences on the trunk of the phylogenetic tree tend to have mutations that our experiments measure to be more beneficial for viral growth than mutations that occur in side-branch HA sequences.
The fact that our experiments can help distinguish between trunk and side-branch lineages suggests that they might inform evolutionary forecasting.
In their landmark paper introducing predictive viral fitness models that accounted for both antigenic and non-antigenic mutations, {\L}uksza and L{\"a}ssig~\cite{luksza2014predictive} noted that the models could in principle be improved by integrating ``diverse genotypic and phenotypic data'' that more realistically represented the effects of specific mutations.
Our work suggests that deep mutational scanning can provide such data.

@@

To parse trunk mutations from side branch mutations, we defined a set of recent nodes sampled on or after January 1, 2017, and traced these nodes back to their most recent common ancestor.
We defined all branches ancestral to this most recent common ancestor as the trunk.
All nodes that do not have descendants isolated after 2017 were defined as side branch nodes.
Nodes descended from the most recent common ancestor were defined as unresolved, as it is still unclear whether these nodes will ultimately be on the trunk or side branches.
The phylogenetic tree is available as a JSON file on \comment{link to repository}.
The tree was visualized using baltic (https://github.com/blab/baltic).
